Validation of Visual Statistical Inference, Applied to Linear Models (supplementary materials)

### 2013-04-23 supplementary.tex
supplementary section 6.7 and 6.8 cleaned

The materials in this document supplement the information presented in the manuscript ``Validation of Visual Statistical Inference, Applied to Linear Models''. Section \ref{sec:proof_lemma} presents the proof of the Lemma \ref{p-lemma} in the manuscript. Section \ref{sec:lineup_selection} describes how the lineups are presented to the subjects for evaluation.  The data cleaning process and how that may affect the result are discussed in Section \ref{sec:data_cleaning} which supplements Section \ref{p-sec:data_cleaning} of the manuscript. The detailed discussion on how much the choice of null plots affect is included in Section \ref{sec:null_choice} which supplement the information in Section \ref{p-sec:null_choice} of the manuscript. Section \ref{sec:typeIII_error} includes the detailed discussion about the Type-III error supplementing Section  \ref{p-sec:typeIII_error}.

@@

\section{Selection of Lineups for each subject} \label{sec:lineup_selection}
Table \ref{tbl:dist_lineup1} shows how the lineups are presented to the subjects for evaluation for the simulation experiment with categorical covariate described in Section \ref{p-sec:category} of the manuscript . The lineups are divided into four groups such as easy, medium, hard and mixed based on the parameter combinations shown in the table. The number of lineups randomly chosen for each difficulty level is shown in the table as well. Note that every subject saw at least one easy lineup in the pool of 10 lineups shown. Some subjects saw more than one easy lineup that comes from the mixed group. The number of total evaluations required for each lineup is also shown in the table and a total of 300 subjects would provide a total of 3000 evaluations that should roughly satisfy that plan.

@@

Following six approaches have been examined while cleaning the data.  Table \ref{tbl:summary} displays number of subjects and their total evaluations in each screening criteria. The criteria 1 includes all the participants hence showing number of total subjects participated in the survey for all the experiments. For our paper we used criteria six to clean the data and the results presented in the paper are based on the cleaned data.

@@

 Visual inference falls into the same framework as randomization tests, where the statistics from the data are compared with those from null data. 
 But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference, generally, between  plots is almost impossible. However,  in this controlled setting we can use $p$-values of the test statistic calculated on the data used in each plot as a proxy for similarity of structure between the plots.

@@

A little known error amongst statisticians is what was coined as Type III error in \citet{mosteller:48}. This occurs when the null hypothesis is correctly rejected but for the wrong reason. Experiment 3 is prone to this type of error. Observers were asked to identify the plot that had the largest slope, either positive or negative. But the actual data plot had a cluster of points, the contamination that forced the conventional test to fail to see any trend. For the human eye this cluster of points is  as visible as the association between the remaining points, leading the observer to detect the actual data plot by looking for the cluster instead of the slope. This would be considered to be a Type III error because it is not what was requested, but  still gives the correct answer. This is not a problem with the lineup protocol, because we want observers to react to anything that sets the the actual data plot apart. Here, in the controlled but simple setting, it is of interest to see what observers cued to in the actual data plot.

Figure \ref{p-fig:power_loess_effect} in the manuscript indicates that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons observers gave for their choices. Observers were provided with four possible reasons to use for their choice:

@@

For visual inference, making a Type III, is not actually a problem. It is only a possibility in this experiment because we are working with known structure. In the real setting, we are excited to see observers detecting the actual data plot, and curious about how they detect it, with all possible reasons encapsulated in the alternative hypothesis. This does show the importance of the qualitative answers that observers give for their choices.

>>>>

The materials in this document supplement the information presented in the manuscript ``Validation of Visual Statistical Inference, Applied to Linear Models''. Section \ref{sec:proof_lemma} presents the proof of the Lemma \ref{p-lemma} in the manuscript. Section \ref{sec:lineup_selection} describes how the lineups are presented to the subjects for evaluation.  The data cleaning process and how that may affect the result are discussed in Section \ref{sec:data_cleaning} which supplements Section \ref{p-sec:data_cleaning} of the manuscript. The detailed discussion on how much the choice of null plots affect is included in Section \ref{sec:null_choice} which supplement the information in Section \ref{p-sec:null_choice} of the manuscript. Section \ref{sec:typeIII_error} includes the detailed discussion about the Type-III error supplementing Section  \ref{p-sec:typeIII_error} of the manuscript.

@@

\section{Selection of Lineups for each subject} \label{sec:lineup_selection}
Table \ref{tbl:dist_lineup1} shows how the lineups are presented to the subjects for evaluation for the simulation experiment with categorical covariate described in Section \ref{p-sec:category} of the manuscript. The lineups are divided into four groups such as easy, medium, hard and mixed based on the parameter combinations shown in the table. The number of lineups randomly chosen for each difficulty level is shown in the table as well. Note that every subject saw at least one easy lineup in the pool of 10 lineups shown. Some subjects saw more than one easy lineup that comes from the mixed group. The number of total evaluations required for each lineup is also shown in the table. For example,  a lineup with sample size 100, standard error 5 and slope parameter 3 requires 203 evaluations so that the proportion correct can be estimated with margin of error 0.05 as described in the manuscript.  A total of 300 subjects would provide a total of 3000 evaluations that should roughly satisfy that plan.  We see in Table \ref{tbl:summary} that the total number of subjects actually participated in experiment 1 is 520 which is much higher than 300. The goal was to get total number of subject close to 300.

@@

Following six approaches have been examined while cleaning the data.  Table \ref{tbl:summary} displays number of subjects and their total evaluations in each screening criteria. The criteria 1 includes all the participants hence showing number of total subjects participated in the survey for all the experiments. As per the experimental design we always show at least one easy lineup and that lineup is used to process the payment of the subject. Thus, for our paper we used criteria six to clean the data and the results presented in the paper are based on the cleaned data.

@@

It is discussed in the Section \ref{p-sec:null_choice} of the manuscript that $p$-values can be used to quantify the similarity of the visual pattern in the plots used for the simulation experiments. Based on this, we explore more details on how much the null plots affect the choice made by the subjects.

@@

%A little known error amongst statisticians is what was coined as Type III error in \citet{mosteller:48}. This occurs when the null hypothesis is correctly rejected but for the wrong reason. Experiment 3 is prone to this type of error. Observers were asked to identify the plot that had the largest slope, either positive or negative. But the actual data plot had a cluster of points, the contamination that forced the conventional test to fail to see any trend. For the human eye this cluster of points is  as visible as the association between the remaining points, leading the observer to detect the actual data plot by looking for the cluster instead of the slope. This would be considered to be a Type III error because it is not what was requested, but  still gives the correct answer. This is not a problem with the lineup protocol, because we want observers to react to anything that sets the the actual data plot apart. Here, in the controlled but simple setting, it is of interest to see what observers cued to in the actual data plot.

Figure \ref{p-fig:power_loess_effect} in the manuscript indicates that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons participants gave for their choices. Participants were provided with four possible reasons to use for their choice:

@@

%For visual inference, making a Type III, is not actually a problem. It is only a possibility in this experiment because we are working with known structure. In the real setting, we are excited to see observers detecting the actual data plot, and curious about how they detect it, with all possible reasons encapsulated in the alternative hypothesis. This does show the importance of the qualitative answers that observers give for their choices.
